Statistical Guarantees for Local Spectral Clustering on Random Neighborhood Graphs

### 2021-12-23 jmlr/final/19-1016.tex

\noindent The proofs of our major theorems largely consist of (at most) three modular parts.
\begin{enumerate}
	\item \textbf{Fixed graph results.} Results which hold with respect to an arbitrary graph $G$, and are stated with respect to functionals (i.e. normalized cut, conductance, and local spread) of $G$;
	\item \textbf{Sample-to-population results.} For the specific choice $G = G_{n,r}$, results relating the aforementioned functionals to their population analogues. 
	\item \textbf{Bounds on population functionals.} (In the case of density clustering only.) When the candidate cluster is a $\lambda$-density cluster, bounds on population functionals as a function of $\lambda$, as well as the other relevant parameters introduced in Section~\ref{sec:ppr_density_cluster}.
\end{enumerate}

>>>>

\documentclass[twoside,11pt]{article}

%%%%%%%%% Packages %%%%%%%%%%%
\usepackage[export]{adjustbox}
\usepackage{subcaption}
\usepackage[abbrvbib]{jmlr2e}

%% Essentials
\usepackage{mathtools}
\usepackage{float}
\usepackage{xcolor}
\usepackage{algorithm,algorithmic}
\usepackage[shortlabels]{enumitem}     
\usepackage{bm}

%% Beautify
\usepackage{xspace}         % better line spacing
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\hypersetup{
	colorlinks,
	citecolor=blue,
	urlcolor=blue,
	linkcolor=blue,
	bookmarks=false
}

%%%%%%%%% Macros %%%%%%%%%%%
\DeclareFontFamily{U}{mathx}{\hyphenchar\font45}
\DeclareFontShape{U}{mathx}{m}{n}{<-> mathx10}{}
\DeclareSymbolFont{mathx}{U}{mathx}{m}{n}
\DeclareMathAccent{\wb}{0}{mathx}{"73}

\newcommand{\eqdist}{\ensuremath{\stackrel{d}{=}}}
\newcommand{\Graph}{\mathcal{G}}
\newcommand{\Reals}{\mathbb{R}}
\newcommand{\Identity}{\mathbb{I}}
\newcommand{\Xsetistiid}{\overset{\text{i.i.d}}{\sim}}
\newcommand{\convprob}{\overset{p}{\to}}
\newcommand{\convdist}{\overset{w}{\to}}
\newcommand{\Expect}[1]{\mathbb{E}\left[ #1 \right]}
\newcommand{\Risk}[2][P]{\mathcal{R}_{#1}\left[ #2 \right]}
\newcommand{\Prob}[1]{\mathbb{P}\left( #1 \right)}
\newcommand{\iset}{\mathbf{i}}
\newcommand{\jset}{\mathbf{j}}
\newcommand{\myexp}[1]{\exp \{ #1 \}}
\newcommand{\abs}[1]{\left \lvert #1 \right \rvert}
\newcommand{\restr}[2]{\ensuremath{\left.#1\right|_{#2}}}
\newcommand{\ext}[1]{\widetilde{#1}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\seq}[1]{\set{#1}_{n \in \N}}
\newcommand{\Xsetotp}[2]{\langle #1, #2 \rangle}
\newcommand{\floor}[1]{\left\lfloor #1 \right\rfloor}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\Cov}{\mathrm{Cov}}
\newcommand{\Xsetiam}{\mathrm{diam}}

\newcommand{\emC}{C_n}
\newcommand{\emCpr}{C'_n}
\newcommand{\emCthick}{C^{\sigma}_n}
\newcommand{\emCprthick}{C'^{\sigma}_n}
\newcommand{\emS}{S^{\sigma}_n}
\newcommand{\estC}{\widehat{C}_n}
\newcommand{\hC}{\hat{C^{\sigma}_n}}
\newcommand{\spansp}{\mathrm{span}~}
\newcommand{\1}{\mathbf{1}}

\newcommand{\Linv}{L^{\Xsetagger}}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator*{\argmax}{argmax}

\newcommand{\emF}{\mathbb{F}_n}
\newcommand{\emG}{\mathbb{G}_n}
\newcommand{\emP}{\mathbb{P}_n}
\newcommand{\F}{\mathcal{F}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\R}{\mathcal{R}}
\newcommand{\Rd}{\Reals^d}
\newcommand{\Nbb}{\mathbb{N}}

%%% Vectors
\newcommand{\thetast}{\theta^{\star}}
\newcommand{\betap}{\beta^{(p)}}
\newcommand{\betaq}{\beta^{(q)}}
\newcommand{\vardeltapq}{\varDelta^{(p,q)}}

%%% Norms
\DeclarePairedDelimiterX{\norm}[1]{\lVert}{\rVert}{#1}
\DeclarePairedDelimiterX{\seminorm}[1]{\lvert}{\rvert}{#1}

%%% Matrices
\newcommand{\X}{X} % no bold
\newcommand{\Y}{Y} % no bold
\newcommand{\Z}{Z} % no bold
\newcommand{\Lgrid}{L_{\grid}}
\newcommand{\Xsetgrid}{D_{\grid}}
\newcommand{\Linvgrid}{L_{\grid}^{\Xsetagger}}
\newcommand{\Lap}{{\bf L}}
\newcommand{\NLap}{{\bf N}}
\newcommand{\PLap}{{\bf P}}

%%% Sets and classes
\newcommand{\Xset}{\mathcal{X}}
\newcommand{\Sset}{\mathcal{S}}
\newcommand{\Hclass}{\mathcal{H}}
\newcommand{\Pclass}{\mathcal{P}}
\newcommand{\Leb}{L}
\newcommand{\mc}[1]{\mathcal{#1}}
\newcommand{\mbb}[1]{\mathbb{#1}}

%%% Distributions and related quantities
\newcommand{\Pbb}{\mathbb{P}}
\newcommand{\Ubb}{\mathbb{U}}
\newcommand{\Ebb}{\mathbb{E}}
\newcommand{\Qbb}{\mathbb{Q}}
\newcommand{\Ibb}{\mathbb{I}}

%%% Operators
\newcommand{\Tadj}{T^{\star}}
\newcommand{\Xsetive}{\mathrm{div}}
\newcommand{\Xsetif}{\mathop{}\!\mathrm{d}}
\newcommand{\gradient}{\mathcal{D}}
\newcommand{\Hessian}{\mathcal{D}^2}
\newcommand{\dotp}[2]{\langle #1, #2 \rangle}
\newcommand{\Dotp}[2]{\Bigl\langle #1, #2 \Bigr\rangle}

%%% Misc
\newcommand{\grid}{\mathrm{grid}}
\newcommand{\critr}{R_n}
\newcommand{\Xsetx}{\,dx}
\newcommand{\Xsety}{\,dy}
\newcommand{\Xsetr}{\,dr}
\newcommand{\Xsetxpr}{\,dx'}
\newcommand{\Xsetypr}{\,dy'}
\newcommand{\wt}[1]{\widetilde{#1}}
\newcommand{\wh}[1]{\widehat{#1}}
\newcommand{\ol}[1]{\overline{#1}}
\newcommand{\spec}{\mathrm{spec}}
\newcommand{\LE}{\mathrm{LE}}
\newcommand{\LS}{\mathrm{LS}}
\newcommand{\OS}{\mathrm{OS}}
\newcommand{\PLS}{\mathrm{PLS}}
\newcommand{\dist}{\mathrm{dist}}
\newcommand{\vol}{\mathrm{vol}}
\newcommand{\cut}{\mathrm{cut}}
\newcommand{\diam}{\mathrm{diam}}
\newcommand{\qed}{\hfill $\blacksquare$}

%%% Order of magnitude
\newcommand{\soom}{\sim}

%%%%%%%%% Headings %%%%%%%%%%%

\usepackage{lastpage}
\jmlrheading{22}{2021}{1-\pageref{LastPage}}{12/19; Revised
	9/21}{11/21}{19-1016}{Alden Green, Sivaraman Balakrishnan, and Ryan J. Tibshirani}
\ShortHeadings{PPR Clustering on Neighborhood Graphs}{Green, Balakrishnan and Tibshirani}
\firstpageno{1}

\begin{document}

\title{Statistical Guarantees for Local Spectral Clustering on Random Neighborhood Graphs}

\author{\name Alden Green \email ajgreen@stat.cmu.edu \\
	\addr Department of Statistics and Data Science\\
	Carnegie Mellon University\\
	Pittsburgh, PA 15213
	\AND
	\name Sivaraman Balakrishnan \email siva@stat.cmu.edu \\
	\addr Department of Statistics and Data Science\\
	Machine Learning Department\\
	Carnegie Mellon University\\
	Pittsburgh, PA 15213
	\AND
	Ryan J. Tibshirani \email ryantibs@stat.cmu.edu \\
	\addr Department of Statistics and Data Science\\
	Machine Learning Department\\
	Carnegie Mellon University\\
	Pittsburgh, PA 15213}

\editor{Ingo Steinwart}

\maketitle

\begin{abstract}% <- trailing % for backward compatibility of style file
	We study the Personalized PageRank (PPR) algorithm, a local spectral method for clustering, which extracts clusters using locally-biased random walks around a given seed node.  In contrast to previous work, we adopt a classical statistical learning setup, where we obtain samples from an unknown nonparametric distribution, and aim to identify sufficiently salient clusters.  We introduce a trio of population-level functionals---the \emph{normalized cut}, \emph{conductance}, and \emph{local spread}, analogous to graph-based functionals of the same name---and prove that PPR, run on a neighborhood graph, recovers clusters with small population normalized cut and large conductance and local spread. We apply our general theory to establish that PPR identifies connected regions of high density (density clusters) that satisfy a set of natural geometric conditions. We also show a converse result, that PPR can fail to recover geometrically poorly-conditioned density clusters, even asymptotically. Finally, we provide empirical support for our theory.
\end{abstract}

\begin{keywords}
	graphs, spectral clustering, Personalized PageRank, density clustering, unsupervised learning
\end{keywords}

\section{Introduction}
In this paper, we consider the problem of clustering: splitting a given data set into groups that satisfy some notion of within-group similarity and between-group difference.  Our particular focus is on spectral clustering methods, a family of powerful nonparametric clustering algorithms. Roughly speaking, a spectral algorithm first constructs a geometric graph $G$, where vertices correspond to samples, and edges correspond to proximities between samples. The algorithm then estimates a feature embedding based on a (suitable) Laplacian matrix of $G$, and applies a simple clustering technique (like $k$-means clustering) in the embedded feature space.

When applied to geometric graphs built from a large number of samples, global spectral clustering methods can be computationally cumbersome and insensitive to the local geometry of the underlying distribution \citep{leskovec2010,mahoney2012}.  This has led to increased interest in \emph{local} spectral clustering algorithms, which leverage locally-biased spectra computed using random walks around some user-specified seed node.  A popular local clustering algorithm is the Personalized PageRank (PPR) algorithm, first introduced by \citet{haveliwala2003}, and then further developed by several others \citep{spielman2011,spielman2014,andersen2006,mahoney2012,zhu2013}.

Local spectral clustering techniques have been practically very successful \citep{leskovec2010,andersen2012,gleich2012,mahoney2012,wu2012}, which has led many authors to develop supporting theory \citep{spielman2013,andersen2009,gharan2012,zhu2013} that gives worst-case guarantees on traditional graph-theoretic notions of cluster quality (such as normalized cut and conductance). In contrast, in this paper we adopt a classical statistical viewpoint, and examine what the output of local clustering on a data set reveals about the underlying density $f$ of the samples. We establish conditions on $f$ under which PPR, when appropriately tuned and initialized inside a candidate cluster $\mc{C} \subseteq \Rd$, will approximately recover this candidate cluster. We pay special attention to the case where $\mc{C}$ is a \emph{density cluster} of $f$---defined as a connected component of the upper level set $\{x \in \Rd : f(x) \geq \lambda\}$ for some $\lambda > 0$---and show precisely how PPR accounts for both geometry and density in estimating a cluster.

Before giving a more detailed overview of our main results, we formally define PPR on a neighborhood graph, review some of the aforementioned worst-case guarantees, and introduce the population-level functionals that govern the behavior of local clustering in our statistical context.

\subsection{PPR Clustering}
\label{subsec:ppr_clustering}
We start by reviewing the PPR clustering algorithm. Let $G = (V,E)$ be an undirected, unweighted, and connected graph. We denote by $A \in \Reals^{n \times n}$ the adjacency matrix of $G$, with entries $A_{uv} = 1$ if $(u,v) \in E$ and $0$ otherwise.  We also denote by $D$ the diagonal degree matrix, with entries $D_{uu} := \sum_{v \in V} A_{uv}$, and by $I$ the $n \times n$ identity matrix. The PPR vector $p_v = p(v,\alpha;G)$ is defined with respect to a given seed node $v \in V$ and a teleportation parameter $\alpha \in [0,1]$, as the solution of the following linear system:
\begin{equation}
\label{eqn:ppr_vector}
p_v = \alpha e_{v} + (1 - \alpha) p_v W,
\end{equation}
where $W = (I + D^{-1}A)/2$ is the lazy random walk matrix over $G$ and $e_{v}$ is the indicator vector for node $v$ (that has a 1 in position $v$ and 0 elsewhere).

In practice, exactly solving the system of equations~\eqref{eqn:ppr_vector} to compute the PPR vector may be too computationally expensive. To address this limitation, \citet{andersen2006} introduced the \emph{$\varepsilon$-approximate} PPR vector (aPPR), which we will denote by \smash{$p_v^{(\varepsilon)}$}. We refer the curious reader to \citet{andersen2006} for a formal algorithmic definition of the aPPR vector, and limit ourselves to highlighting a few salient points: the aPPR vector can be computed in order $\mc{O}(1/(\varepsilon\alpha))$ time, while satisfying the following uniform error bound: 
\begin{equation}
\label{eqn:appr_error}
p_v(u) - \varepsilon D_{uu}\leq p_v^{(\varepsilon)}(u) \leq p_v(u), 
\quad \text{for all $u \in V$}.
\end{equation}

Once $p_v$ or \smash{$p_v^{(\varepsilon)}$} is computed, the cluster estimate \smash{$\wh{C}$} is chosen by taking a particular sweep cut. For a given level $\beta > 0$, the \emph{$\beta$-sweep cut} of $p_v = (p_v(u))_{u \in V}$ is 
\begin{equation}
\label{eqn:sweep_cuts}
S_{\beta,v} := \set{u \in V: \frac{p_v(u)}{D_{uu}} > \beta}.
\end{equation}
To define \smash{$\wh{C}$}, one computes $S_{\beta,v}$ over all \smash{$\beta \in (L, U)$} (where the range $(L,U)$ is user-specified), and then outputs the cluster estimate \smash{$\wh{C} = S_{\beta^*,v}$} with minimum normalized cut. For a set $C \subseteq V$ with complement $C^c = V \!\setminus\! C$, the \emph{cut} and \emph{volume} are respectively,
\begin{equation}
\label{eqn:cut_volume}
\cut(C;G) := \sum_{u \in C} \sum_{v \in C^c}
\1\{(u,v) \in E\},~~ \vol(C; G) := \sum_{u \in C}  \sum_{v \in V} \1\{(u,v) \in E\},
\end{equation}
and the \emph{normalized cut} of $C$ is
\begin{equation}
\label{eqn:normalized_cut}
\Phi(C; G) := \frac{\cut(C;G)}{\min \set{\vol(C; G), \vol(C^c; G)}}.
\end{equation}

\subsection{Worst-Case Guarantees for PPR Clustering}
As mentioned, most analyses of local clustering have focused on worst-case guarantees, defined with respect to functionals of an a priori fixed graph $G = (V,E)$.  For instance, \cite{andersen2006} analyze the normalized cut of the cluster estimate $\wh{C}$ output by PPR, showing that when PPR is appropriately seeded within a candidate cluster $C \subseteq V$, the normalized cut $\Phi(\wh{C};G)$ is upper bounded by (a constant times)  $\sqrt{\Phi(C;G)}$. \cite{zhu2013} build on this: they introduce a second functional, the \emph{conductance} $\Psi(G)$, defined as 
\begin{equation}
\label{eqn:conductance}
\Psi(G) := \min_{S \subseteq V} \Phi(S;G),
\end{equation}
and show that if $\Phi(C;G)$ is much smaller than $\Psi(G[C])^2$---where $G[C] = (C,E \cap (C \times C))$ is the subgraph of $G$ induced by $C$--- then (in addition to having a small normalized cut) the cluster estimate $\wh{C}$ approximately recovers $C$. Our own analysis builds on that of~\cite{zhu2013}, and we present a more detailed summary of their results in Section~\ref{sec:ub_symmetric_set_difference}. For now, we merely reiterate that the conclusions of \citet{andersen2006,zhu2013} cannot be straightforwardly applied to our setting, where the input data are random samples $\{x_1,\ldots,x_n\}$ drawn from a distribution $\Pbb$, the graph $G$ is a random neighborhood graph formed from the samples, and the candidate cluster is a set $\mc{C} \subseteq \Rd$.\footnote{Throughout, we use calligraphic notation to refer to subsets of $\Rd$.}

\subsection{PPR on a Neighborhood Graph}
\label{subsec:ppr_neighborhood_graph}
We now formally describe the statistical setting in which we operate, as well as the method we will study: PPR on a neighborhood graph. Let $X = \{x_1,\ldots, x_n\}$ be samples drawn i.i.d.\ from a distribution $\Pbb$ on $\Rd$. We will assume throughout that $\Pbb$ has a density $f$ with respect to the Lebesgue measure $\nu$ on $\Rd$ . For a radius $r > 0$, we define $G_{n,r}=(V,E)$ to be the \emph{$r$-neighborhood graph} of $X$, an unweighted, undirected graph with vertices $V=X$, and an edge $(x_i,x_j) \in E$ if and only if $i \neq j$ and $\|x_i - x_j\| \leq r$, where $\|\cdot\|$ is the Euclidean norm. Once the neighborhood graph $G_{n,r}$ is formed, the PPR vector $p_v$ is then computed over $G_{n,r}$, with a resulting cluster estimate \smash{$\wh{C} \subseteq X$}. The precise algorithm is summarized in Algorithm~\ref{alg:ppr}.

\begin{algorithm}
	\caption{PPR on a neighborhood graph}
	\label{alg:ppr}	
	{\bfseries Input:} data $X=\{x_1,\ldots,x_n\}$, radius $r > 0$, teleportation
	parameter $\alpha \in [0,1]$, seed $v \in X$, sweep cut range $(L,U)$. \\     
	{\bfseries Output:} cluster estimate $\wh{C} \subseteq V$.
	\begin{algorithmic}[1]
		\STATE Form the neighborhood graph $G_{n,r}$.
		\STATE Compute the PPR vector $p_v=p(v, \alpha; G_{n,r})$ as in
		\eqref{eqn:ppr_vector}.  
		\STATE Compute sweep cuts $S_{\beta}$ as in \eqref{eqn:sweep_cuts}, for each $\beta \in (L,U)$.\footnotemark 
		\STATE Return \smash{$\wh{C} = S_{\beta^*}$}, where  
		$$
		\beta^* = \argmin_{\beta \in (L,U)}~ \Phi(S_{\beta}; G_{n,r}).
		$$
	\end{algorithmic}
\end{algorithm}
\footnotetext{Technically speaking, for each $\beta \in (L,U) \cap \{p_v(u)/D_{uu}: u \in V\}$.}

\subsection{Cluster Accuracy}
We need a metric to assess the accuracy with which $\wh{C}$ estimates the candidate cluster $\mc{C}$. One commonly used metric is the misclassification error, i.e., the size of the symmetric set difference between \smash{$\wh{C}$} and the empirical cluster $\mc{C}[X] = \mc{C} \cap X$ \citep{korostelev1993,polonik1995,rigollet2009}. We will consider a related metric, the volume of the symmetric set difference, which weights misclassified points according to their degree in $G_{n,r}$. To keep things simple, for a given set $S \subseteq X$ we write $\vol_{n,r}(S) := \vol(S;G_{n,r})$.

\begin{definition}
	\label{def:volume_symmetric_set_difference}
	For an estimator \smash{$\wh{C} \subseteq X$} and a set $\mc{C} \subseteq \Rd$, their symmetric set difference is  
	\begin{equation*}
	\wh{C} \vartriangle \mc{C}[X] :=
	\bigl(\wh{C} \setminus \mc{C}[X]\bigr) \cup
	\bigl(\mc{C}[X] \setminus \wh{C}\bigr).
	\end{equation*}
	Furthermore, we denote the volume of the symmetric set difference by 
	$$
	\Delta(\wh{C}, \mc{C}[X]) := \vol_{n,r}(\wh{C} \vartriangle \mc{C}[X]). 
	$$
\end{definition}

\subsection{Population Normalized Cut, Conductance, and Local Spread}
Next we define three population-level functionals of $\mc{C}$---the normalized cut $\Phi_{\Pbb,r}(\mc{C})$, conductance $\Psi_{\Pbb,r}(\mc{C})$, and the local spread $s_{\Pbb,r}(\mc{C})$---which we will use to upper bound the volume of the symmetric set difference \smash{$\Delta(\wh{C},\mc{C}[X])$} with high probability.

Let the population-level \emph{cut} of $\mc{C}$ be the expectation (up to a rescaling) of $\cut_{n,r}(\mc{C}[X]) := \cut(\mc{C}[X]; G_{n,r})$,  and likewise let the population-level \emph{volume} of $\mc{C}$ be the expectation (up to a rescaling) of $\vol_{n,r}(\mc{C}[X]) := \vol(\mc{C}[X]; G_{n,r})$; i.e., let
\begin{equation*}
\mathrm{cut}_{\Pbb,r}(\mc{C}) := \int_{\mc{C}} \int_{\mc{C}^c} \1\{\|x - y\| \leq r\} \,d\Pbb(y) \,d\Pbb(x),~~ \mathrm{vol}_{\Pbb,r}(\mc{C}) := \int_{\mc{C}} \int_{\Rd} \1\{\|x - y\| \leq r\} \,d\Pbb(y) \,d\Pbb(x),
\end{equation*}
where $\mc{C}^c := \Rd \!\setminus\! \mc{C}$. Also let $\deg_{\Pbb,r}(x) := \int_{\Rd} \1\{\|y - x\| \leq r\} \,d\Pbb(y)$ be the expected degree of $x$ in $G_{n,r}$. 
\begin{definition}[Population normalized cut]
	For a set $\mc{C} \subseteq \Rd$, distribution $\Pbb$, and radius $r > 0$, the \emph{population normalized cut} is
	\begin{equation}
	\label{eqn:population_normalized_cut}
	\Phi_{\Pbb,r}(\mc{C}) := \frac{\mathrm{cut}_{\Pbb,r}(\mc{C})}{\min\{\mathrm{vol}_{\Pbb,r}(\mc{C}), \mathrm{vol}_{\Pbb,r}(\mc{C}^c)\}}.
	\end{equation}
\end{definition}

Let \smash{$\wt{\Pbb}(\cdot) = \Pbb(\cdot|x \in \mc{C})$} be the conditional distribution of $x$, i.e., let \smash{$\wt{\Pbb}(\mc{S}) = \wt{\Pbb}(\mc{S} \cap \mc{C})/\wt{\Pbb}(\mc{C})$} for measurable sets $\mc{S} \subseteq \Rd$.

\begin{definition}[Population conductance]
	For a set $\mc{C} \subseteq \Rd$, distribution $\Pbb$ and radius $r > 0$, the \emph{population conductance} is
	\begin{equation}
	\label{eqn:population_conductance}
	\Psi_{\Pbb,r}(\mc{C}) = \inf_{\mc{S} \subseteq \mc{C}} \Phi_{\wt{\Pbb},r}(\mc{S}).
	\end{equation}
\end{definition}

\begin{definition}[Population local spread]
	For a set $\mc{C} \subseteq \Rd$, distribution $\Pbb$ and radius $r > 0$, the \emph{population local spread} is
	\begin{equation}
	\label{eqn:local_spread}
	s_{\Pbb,r}(\mc{C}) := \min_{x \in \mc{C}} \biggl\{\frac{\bigl(\deg_{\wt{\Pbb},r}(x)\bigr)^2}{\vol_{\wt{\Pbb},r}(\mc{C})} \biggr\},
	\end{equation}
\end{definition}

It is quite natural that $\Phi_{\Pbb,r}(\mc{C})$ and $\Psi_{\Pbb,r}(\mc{C})$ should help quantify the role geometry plays in local spectral clustering. Indeed, these functionals are the population-level analogues of the empirical quantities $\Phi_{n,r}(\mc{C}[X]) := \Phi(\mc{C}[X];G_{n,r})$ and $\Psi_{n,r}(\mc{C}[X]) := \Psi(G_{n,r}\bigl[\mc{C}[X]\bigr])$, and as we have already mentioned, these empirical quantities can be used to upper bound the volume of the symmetric set difference. For this reason, similar population-level functionals are used by \cite{shi2009,schiebinger2015,garciatrillos19} in the analysis of \emph{global} spectral clustering in a statistical context. We will comment more on the relationship between these works and our own results in Section~\ref{subsec:related_work}.

The role played by $s_{\Pbb,r}(\mc{C})$ is somewhat less obvious. For now, we mention only that it plays an essential part in obtaining tight bounds on the mixing time of a particular random walk that is closely related to the PPR vector, and defer further discussion until later in Section~\ref{sec:ub_symmetric_set_difference}.

\subsection{Main Results}
We now informally state our two main upper bounds, regarding the recovery of a generic cluster $\mc{C}$, and a density cluster $\mc{C}_{\lambda}$. Theorem~\ref{thm:volume_ssd_ub_informal} informally summarizes the first of our main results (formally stated in Theorem~\ref{thm:volume_ssd_ub}) regarding the recovery of a generic cluster $\mc{C}$.
\begin{theorem}[Informal]
	\label{thm:volume_ssd_ub_informal}
	If $\mc{C} \subseteq \Rd$ and $\Pbb$ satisfy appropriate regularity conditions, and Algorithm~\ref{alg:ppr} is initialized properly with respect to $\mc{C}$, then for all sufficiently large $n$, with high probability it holds that
	\begin{equation*}
	\frac{\Delta(\wh{C},\mc{C}[X])}{\vol_{n,r}(\mc{C}[X])} \leq c \cdot\Phi_{\Pbb,r}(\mc{C}) \cdot  \biggl(\frac{\log(1/s_{\Pbb,r}(\mc{C}))}{\Psi_{\Pbb,r}(\mc{C})}\biggr)^2.
	\end{equation*}
\end{theorem}

(Above, and throughout, $c$ stands for a universal constant that may change from line to line.) Put more succinctly, we find that \smash{$\Delta(\wh{C},\mc{C}[X])$} is small when $\Phi_{\Pbb,r}(\mc{C})$ is small relative to $(\Psi_{\Pbb,r}(\mc{C})/\log(1/s_{\Pbb,r}(\mc{C})))^2$. To the best of our knowledge, this gives the first population-level guarantees for local clustering in the nonparametric statistical context.

Next, Theorem~\ref{thm:density_cluster_volume_ssd_ub_informal} informally summarizes the second of our main results (formally stated in Theorem~\ref{thm:density_cluster_volume_ssd_ub}) regarding the recovery of a $\lambda$-density cluster $\mc{C}_{\lambda}$ by PPR. For reasons that we explain later in Section~\ref{sec:ppr_density_cluster}, our cluster recovery statement will actually be with respect to the $\sigma$-thickened set $\mc{C}_{\lambda,\sigma} := \{x \in \Rd: \mathrm{dist}(x,\mc{C}_{\lambda}) < \sigma\}$, for a given $\sigma > 0$. The upper bound we establish is a function of various parameters that measure the conditioning of both the density cluster $\mc{C}_{\lambda,\sigma}$ and density $f$ for recovery by PPR. We assume that $\mc{C}_{\lambda,\sigma}$ is the image of a convex set $\mc{K}$ of finite diameter $\mathrm{diam}(\mc{K}) \leq \rho < \infty$ under a Lipschitz, measure-preserving mapping $g$, with Lipschitz constant $M$. We also assume that $f$ is bounded away from $0$ and $\infty$ on $\mc{C}_{\lambda,\sigma}$:
\begin{equation*}
0 < \lambda_{\sigma} \leq f(x) \leq \Lambda_{\sigma} < \infty~~\text{for all $x \in \mc{C}_{\lambda,\sigma}$},
\end{equation*}
and additionally satisfies the following low-noise condition:
\begin{equation*}
\inf_{y \in \mc{C}_{\lambda,\sigma}} f(y) - f(x) \geq  \theta \cdot \dist(x, \mc{C}_{\lambda,\sigma})^{\gamma}~~\text{for all $x$ such that $0 < \dist(x,\mc{C}_{\lambda,\sigma}) \leq \sigma$.}
\end{equation*}
(Here $\dist(x,\mc{C}) := \inf_{y \in \mc{C}} \|y - x\|$.) 
\begin{theorem}[Informal]
	\label{thm:density_cluster_volume_ssd_ub_informal}
	If $\mc{C}_{\lambda} \subseteq \Rd$ is a $\lambda$-density cluster of a distribution $\Pbb$, which satisfies appropriate regularity conditions, and Algorithm~\ref{alg:ppr} is initialized properly with respect to $\mc{C}_{\lambda,\sigma}$, then for all sufficiently large $n$, with high probability it holds that
	\begin{equation*}
	\frac{\Delta(\wh{C},\mc{C}_{\lambda,\sigma}[X])}{\vol_{n,r}(\mc{C}_{\lambda,\sigma})} \leq c \cdot d^4 \cdot \frac{M^2\rho^2}{\sigma r} \cdot \frac{\Lambda_{\sigma}^2 \lambda (\lambda - \theta \frac{r^{\gamma}}{\gamma + 1})}{\lambda_{\sigma}^4} \cdot \log^2\biggl(\frac{\Lambda_{\sigma}^{2/d} M\rho}{\lambda_{\sigma}^{2/d}2r}\biggr).
	\end{equation*}
\end{theorem}

The above result reveals the separate roles played by geometry and density in the ability of PPR to recover a density cluster. Here $M$, $\rho$, and $\sigma$ capture whether $\mc{C}_{\lambda,\sigma}$ is geometrically well-conditioned (short and fat) or poorly-conditioned (long and thin) for recovery by PPR. Likewise, the parameters $\lambda_{\sigma}, \Lambda_{\sigma}, \gamma$, and $\theta$ measure whether the density $f$ is well-conditioned (approximately uniform over the density cluster, and having thin tails outside of it) or poorly conditioned (vice versa). Theorem~\ref{thm:density_cluster_volume_ssd_ub_informal} says that if the thickened density cluster $\mc{C}_{\lambda,\sigma}$ is geometrically well-conditioned---meaning, $M^2\rho^2/(\sigma r) \approx 1$---and the density $f$ is well-conditioned near $\mc{C}_{\lambda,\sigma}$---meaning, $\Lambda_{\sigma} \approx \lambda \approx \lambda_{\sigma}$ and $\lambda - \theta r^{\gamma}/(\gamma + 1)$ is much less than $\lambda_{\sigma}$---then PPR will approximately recover $\mc{C}_{\lambda,\sigma}$.

\subsection{Related Work}
\label{subsec:related_work}
We now summarize some related work (in addition to the background material already given above), regarding the theory of spectral clustering, and of density cluster recovery.

\subsubsection{Spectral Clustering} 
In the stochastic block model (SBM), arguably one of the simplest models of network formation, edges between nodes independently occur with probability based on a latent community membership. In the SBM, the ability of spectral algorithms to perform clustering---or community detection---is well-understood, dating back to \citet{mcsherry2001} who gives conditions under which the entire community structure can be recovered. In more recent work, \citet{rohe2011} upper bound the fraction of nodes misclassified by a spectral algorithm for the high-dimensional (large number of blocks) SBM, and \citet{lei2015} extend these results to the sparse (low average degree) regime. Relatedly, \citet{clauset08,balakrishnan2011,li2018} analyze the misclassification rate when the block model exhibits some hierarchical structure. The framework we consider, in which nodes correspond to data points sampled from an underlying density, and edges between nodes are formed based on geometric proximity, is quite different than the SBM, and therefore so is our analysis.

In general, the study of spectral algorithms on neighborhood graphs has been focused on establishing asymptotic convergence of eigenvalues and eigenvectors of certain sample objects to the eigenvalues and eigenfunctions of corresponding limiting operators. \citet{koltchinskii2000} establish convergence of spectral projections of the adjacency matrix to a limiting integral operator, with similar results obtained using simplified proofs in \citet{rosasco10}. \citet{vonluxburg2008} studies convergence of eigenvectors of the Laplacian matrix for a neighborhood graph of fixed radius. \citet{belkin07} and \citet{garciatrillos18} extend these results to the regime where the radius $r \to 0$ as $n \to \infty$.

These results are of fundamental importance. However, they remain silent on the following natural question: do the spectra of these continuum operators induce a partition of the sample space which is ``good'' in some sense? \citet{shi2009,schiebinger2015,garciatrillos19,hoffmann2019} address this question, showing that spectral algorithms will recover the latent labels in certain well-conditioned nonparametric mixture models. These works are probably the most similar to our own: the conditioning of these mixture models depends on population-level functionals resembling the population normalized cut and conductance introduced above, and the resulting bounds on the error of spectral clustering are comparable to those we establish in Theorem~\ref{thm:volume_ssd_ub}. However, these results focus on global rather than local methods, and impose global rather than local conditions on $\Pbb$. Moreover, they do not explicitly consider recovery of density clusters, which is an important concern of our work. We comment further on the relationship between our results and these works after Theorem~\ref{thm:volume_ssd_ub}.

\subsubsection{Density Clustering} 
For a given threshold $\lambda \in (0,\infty)$, we denote by \smash{$\mbb{C}_f(\lambda)$} the connected components of the density upper level set $\{x \in \Rd: f(x) \geq \lambda\}$. In the density clustering problem, initiated by~\cite{hartigan1975}, the goal is to recover $\mbb{C}_{f}(\lambda)$. By now, density clustering, and the related problem of level-set estimation, have been thoroughly studied. For instance, \citet{polonik1995,rigollet2009, rinaldo2010, steinwart2015} study density clustering under the symmetric set difference metric, \citet{tsybakov1997,singh2009,jiang2017} describe minimax optimal level-set and cluster estimators under Hausdorff loss, and \citet{hartigan1981,chaudhuri2010,kpotufe11,balakrishnan2013,steinwart2017,wang2019} consider estimation of the cluster tree $\{\mbb{C}_f(\lambda): \lambda \in (0,\infty)\}$.

We emphasize that our goal is not to improve on these results, nor is it to offer a better algorithm for density clustering. Indeed, seen as a density clustering algorithm, PPR has none of the optimality guarantees found in the aforementioned works. Rather, we hope to better understand the implications of our general theory by applying it within an already well-studied framework. We should also note that since we study a local algorithm, our interest will be in a local version of the density clustering problem, where the goal is to recover a single density cluster $\mc{C}_{\lambda} \in \mbb{C}_f(\lambda)$.
